---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# singleRcapture

<!-- badges: start -->
<!-- badges: end -->

The goal of singleRcapture is to automate single source capture-recapture estimation of population size.

## Installation

You can install the development version of singleRcapture from [GitHub](https://github.com/ncn-foreigners/singleRcapture) with:

``` r
# install.packages("devtools")
devtools::install_github("ncn-foreigners/singleRcapture")
```

## Example

This is a basic example of zero truncated poisson model and zelterman model with netherlands imigrant data with analytic variance:

```{r example}
library(singleRcapture)
ModelPo <- estimate_popsize(formula = capture ~ .,
                            data = netherlandsimmigrant,
                            pop.var = "analytic",
                            model = "ztpoisson",
                            method = "robust")
ModelZl <- estimate_popsize(formula = capture ~ .,
                            data = netherlandsimmigrant,
                            pop.var = "analytic",
                            model = "zelterman",
                            method = "robust")
summary(ModelPo)
```

```{r}
summary(ModelZl)
```

Marginal frequencies and Goodness of fit test:
```{r}
summary(marginalFreq(ModelPo), df = 2)
```

Here is a plot of marginal frequencies with matplot:

```{r plot}
a1 <- marginalFreq(ModelPo)$table[-1]
a2 <- marginalFreq(ModelZl)$table[-1]
a3 <- table(netherlandsimmigrant$capture)
matplot(y = sqrt(cbind(a1, a2, a3)), x = 1:6, ylab = "Square root of Frequencies",
        xlab = "Counts", pch = 21:23, type = "o",
        lty = 1, col = 1:3, main = "Plot of observed and fitted marginal frequencies")
legend("topright",
       legend = c("Poisson model",
                  "Zelterman model",
                  "Observed data"),
       col = 1:3, pch = 21:23)
```
