---
title: "oichao(): Theory, Simulation, and R Code"
author: "singleRcapture playground"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    embed-resources: true
    html-math-method: mathml
execute:
  echo: true
  warning: false
  message: false
---

```{r}
if (dir.exists("/tmp/singleRcapture-lib")) {
  .libPaths(c("/tmp/singleRcapture-lib", .libPaths()))
}

library(singleRcapture)
```

# Overview

This note documents the Poisson-kernel modified Chao family `oichao()`
implemented in `singleRcapture`. The model is intended for single-source
capture-recapture settings where one-inflation is plausible and the usual
generalised Chao estimator may overestimate because it depends directly on
the frequency of count `1`.

The core idea is to replace the usual rare-count regression on counts
`1` and `2` by a regression on counts `2` and `3`. For the Poisson kernel,
the conditional success probability is

$$
\Pr(Y_i = 3 \mid Y_i \in \{2,3\}, x_i) = \frac{\lambda_i}{3 + \lambda_i},
$$

which becomes logistic under the link

$$
\eta_i = \log(\lambda_i / 3).
$$

The resulting hidden-zero contribution is

$$
\frac{p_0(\lambda_i)}{p_2(\lambda_i) + p_3(\lambda_i)}
= \frac{1}{\lambda_i^2 / 2 + \lambda_i^3 / 6},
$$

so the corresponding population size estimator is

$$
\widehat N_{\mathrm{oichao}}
= N_{obs} + \sum_{i:Y_i \in \{2,3\}}
\frac{1}{\widehat \lambda_i^2 / 2 + \widehat \lambda_i^3 / 6}.
$$

In the intercept-only case, the estimator reduces to the modified Chao
closed form

$$
\widehat f_0 = \frac{2}{9}\frac{f_2^3}{f_3^2}.
$$

# Package Example

The new family is available directly through `estimatePopsize()`:

```{r}
fit_oichao <- estimatePopsize(
  capture ~ gender + age,
  data = netherlandsimmigrant,
  model = "oichao",
  method = "IRLS",
  controlMethod = controlMethod(silent = TRUE)
)

popSizeEst(fit_oichao)
```

# Intercept-Only Sanity Check

For `f_2 = 95` and `f_3 = 32`, the fitted intercept implies
$\widehat \lambda = 3f_3 / f_2$, and the regression version matches the
classical modified Chao formula exactly.

```{r}
theory_data <- data.frame(y = c(rep(2, 95), rep(3, 32)))

fit_theory <- estimatePopsize(
  y ~ 1,
  data = theory_data,
  model = "oichao",
  method = "IRLS",
  popVar = "analytic",
  controlMethod = controlMethod(silent = TRUE)
)

lambda_hat <- 3 * exp(unname(coef(fit_theory))[1])
closed_form <- 127 + (2 / 9) * 95 ^ 3 / 32 ^ 2

data.frame(
  lambda_hat = lambda_hat,
  lambda_closed_form = 3 * 32 / 95,
  Nhat_regression = fit_theory$populationSize$pointEstimate,
  Nhat_closed_form = closed_form
)
```

# Simulation Design

The simulation compares the built-in generalised Chao estimator to the
new `oichao()` family under three scenarios:

1. No one-inflation
2. Constant one-inflation
3. Covariate-dependent one-inflation

The data-generating model is

$$
\log(\lambda_i) = 0.1 + 0.7x_{1i} - 0.35x_{2i},
$$

with

$$
x_{1i} \sim \mathrm{Bernoulli}(0.45), \qquad x_{2i} \sim N(0,1),
$$

and observed counts generated by

$$
Y_i =
\begin{cases}
1, & \text{with probability } \omega_i, \\
Y_i^\star, & \text{with probability } 1 - \omega_i,
\end{cases}
\qquad
Y_i^\star \sim \mathrm{Poisson}(\lambda_i).
$$

The exact R code used for the simulation is stored in
`simulate_oichao.R` and included below verbatim.

```{r}
#| echo: false
cat("```r\n")
cat(paste(readLines("simulate_oichao.R"), collapse = "\n"))
cat("\n```\n")
```

# Stored Simulation Summary

The table below is rendered from the saved results in
`modchao_simulation_summary.csv`, which were produced by the simulation
code shown above.

```{r}
summary_tab <- read.csv("modchao_simulation_summary.csv")
knitr::kable(summary_tab, digits = 3)
```

The qualitative pattern matches the intended theory:

- Without one-inflation, the usual generalised Chao estimator is more
  efficient because it uses the more common counts `1` and `2`.
- Under constant or covariate-dependent one-inflation, the generalised
  Chao estimator is strongly upward biased.
- `oichao()` remains close to unbiased in the one-inflated settings,
  which is the main point of the modified Chao construction.

# Reproduce the Summary Table

If you want to rerun the simulation from scratch rather than use the
stored CSV, the following code will reproduce the summary table in R.
It is intentionally not executed during rendering because the full run
uses `B = 500` replications in each scenario.

```{r}
#| eval: false
source("simulate_oichao.R")
```
